Real-time Ray Tracing

### 2020-05-23 portfolio/paper/paper.tex
Progress on report

% Some package I am using. You may not need them
%
\usepackage{rotating}
\usepackage{subfloat}
\usepackage{color}
\usepackage{amsmath}

@@

\paragraph{Spheres} \citep[]{linesphereintersect} Spheres are commonly used in ray tracing to test the ray tracing techniques as the every point on the surface of a sphere has a different normal. The intersection test algorithm is also relatively quick and simple compared to a triangle mesh.

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

% Some package I am using. You may not need them
%
\usepackage{rotating}
\usepackage{subfloat}
\usepackage{color}
\usepackage{amsmath}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{float}
\usepackage[export]{adjustbox}
\usepackage{hyperref}

@@

\paragraph{Spheres} \citep{linesphereintersect} Spheres are commonly used in ray tracing to test the ray tracing techniques as the every point on the surface of a sphere has a different normal. The intersection test algorithm is also relatively quick and simple compared to a triangle mesh.

@@

In most cases, the secondary rays will have to go through all the same scene intersection tests as the primary ray and may even have to perform extra computation for its specific needs when traced.

@@

With the pixel coordinates specified by the kernel global ID's, the normalised ray origins are computed with:

\begin{align}
    N &= \begin{bmatrix}
        (2(x / w) - 1) * r \\
        2(y / h) - 1 \\
        D
    \end{bmatrix}
\end{align}

where $x$ and $y$ are the pixel coordinates, $w$ and $h$ are the image width and height, $r$ is the aspect ratio (height/width), and $D$ is the distance Z-axis offset of the ray origin to the camera position.

Since the camera can move and rotate, primary rays will vary depending on the rotation angle and the camera position. The ray origin can simply be offset by the camera position for translation. For rotation, given the vertical and traversal axis angles, the ray direction is computed by 
$$[ N_x, N_y * cos(a) + N_z * -sin(a), N_z * cos(a) + N_y * sin(a) ]$$ where $N$ is the normalised coordinates of the pixel and $a$ is the traversal axis angle of the camera rotation (pitch). The vertical axis angle (yaw) can then be applied with $$[ O_x * cos(b) + O_z * sin(b), O_y, O_z * cos(b) + O_x * -sin(b) ]$$ where $O$ is the position vector returned by the traversal axis angle formula.

The ray direction is computed by simply taking the normalised vector of the camera position to the ray origin.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{img/backward_ray_tracing.png}
    \caption{Diagram of how primary rays are generated from the camera. \citep{henrikdia}}
\end{figure}

@@

When a ray hits a surface and the surface material is reflective, a child ray for reflection is created. The origin of the reflection ray is the intersection point of its parent ray and the direction of computed by $$I - 2N(I \cdot N)$$ where $I$ is the parent ray direction, $N$ is the surface normal, and the dot represents the dot product. The result of the reflection ray will be mixed into the output result of its parent ray depending on the surface material. If a ray does not intersect with any object or the surface material is not reflective, a reflection ray is not generated. When a reflection ray intersects with an object, it may create a child ray of itself for another reflection ray depending on the bounce limit. The rays can be traced sequentially however the colour results must be computed in a recursive manner to be physically accurate.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{img/reflection_diagram.png}
    \caption{Diagram of how the reflection direction vector is computed. \citep{fitzpatrickreflect}}
\end{figure}

@@

If the ray hits a surface that is not completely opaque, like the reflection ray, a child ray is created for refraction transmission. Similar to reflection, the origin is at the intersection point and the surface normal is used to compute the direction of the refraction ray. The refraction ray will usually be traced through the object that was hit. In order to compute the direction, the refractive indices of the parent ray medium (e.g. air) and object are needed. The following pseudocode is how the refractive direction is computed in the kernel.

\begin{algorithm} [H]
\SetAlgoLined
\KwResult{Direction of the refraction ray: $R_d$}
\KwData{
I = incident direction vector, 
N = surface normal, 
F = previous medium refractive index, 
T = object medium refractive index
}

$n = F / T$ \\
$C = -N \cdot I$ \tcp{Dot product} 
$S = n^2(1 - C^2)$ \\
\tcp{Total internal reflection} 
\uIf{S > 1}{return} 
$T = \sqrt{1 - S}$ \\
$R_d = n \times I + n(nC - T)$

\caption{Algorithm to compute the refraction direction of a given incident direction and surface normal.}
\end{algorithm}

When total internal reflection occurs, the refraction ray is ignored. In most cases, the refraction ray will intersect a surface of the same object (the refraction ray travels through the object) where another refraction ray would be generated, this time from inside the object going away from the object. The $F$ and $T$ values will be opposite in this case compared to the previous refraction computation.

@@

Because rays are traced backwards (from the camera towards light sources), surfaces in shadows need to be calculated differently. In the physical world, there is simply less light coming the surface so it appears darker. However due to the nature of backward ray tracing, a shadow ray is created and traced towards a light source where the surface is in shadow if the shadow ray intersects with any object.

Shadows traced this way will produce very hard shadows where the edges of the shadow will follow very close to the edges of the object which will look unrealistic due to global illumination from light bouncing around in the real world where parts of the shadowed surface will be illuminated by light bouncing off objects nearby. The effects of the hard shadow edges can be mitigated by tracing multiple shadow rays with a slight offset towards the same light source and the strength of the shadow for a particular point on the surface is calculated with $\frac{H}{N}$ where $H$ is the number of shadow rays that have hit an object and $N$ is the total number of shadow rays traced.

This approach of creating shadows is less physically accurate and shows in the image produced as surfaces that are not completely opaque will not let light through like it would in the real world without tracing even more rays and including refraction. With soft shadows, including this feature would hinder performance too much.

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{img/shadow_diagram.png}
    \caption{Diagram of shadow rays \citep{scratchshadow}}
\end{figure}

@@

To produce an image, the rays traced must return information that can be used to compute the result colour of the ray and be mixed to produce the final output colour of a pixel. The information information returned is based on the surface material of the objects that are hit by the rays. The surface material includes information such as diffuse colour, opacity, reflectance, and refractive index of the object medium.

@@

Ray intersection with spheres are simple and quick. A sphere is stored in memory as a C struct containing its position, radius, and material. For better memory efficiency, the material is stored as an index to a separate material buffer as many objects may use the same material. Sphere intersection is calculated with the following pseudocode.

\begin{algorithm}[H]
    \SetAlgoLined
    \KwResult{True if ray intersects sphere, else false}
    \KwData{Ray $R$, sphere position $S_o$, sphere radius $S_r$}

    \tcp{Parametric equation: $at^2 + bt + c = 0$}
    $b = 2(R_d \cdot (R_o - S_o))$ \tcp{Dot product}
    $c = (R_o - S_o)^2 - S_r^2$ \\
    \uIf{$b^2 - 4c < 0$}{return false}
    $d = \sqrt{b^2 - 4c}$ \tcp{Discriminant}
    $T_{min} = (-b - d) / 2$ \\
    $T_{max} = (-b + d) / 2$ \\

    return true

\caption{Procedure used to determine if a ray intersects with a sphere.}
\end{algorithm}

$T_min$ and $T_max$ can be used to find the intersection points of the sphere. The ray does not intersect with the sphere if the discriminant is negative. The normal of the surface is simply the vector from the sphere center to the surface intersection point.

@@

While spheres are simple and quick, they are not the most versatile shape for complex objects. Any 3D model can be represented as a triangular mesh so a ray-triangle intersection test was also implemented. Triangle intersection tests are slower than sphere intersections and 3D models can have a performance crippling number of triangles to test however there are methods to reduce the number of triangles a ray needs to test for intersection.

\begin{algorithm}[H]
    \SetAlgoLined
    \KwResult{True if ray intersects triangle, else false}
    \KwData{Ray $R$, Triangle vertices $V1, V2, V3$}
    $edge_1 = V2 - V1$ \\
    $edge_2 = V3 - V1$ \\
    $h = R_d \times edge_2$ \tcp{Cross product}
    $a = edge_1 \cdot edge_2$ \\
    \tcp{If ray is parallel to triangle}
    \uIf{a = 0}{ return false}
    $f = 1 / a$ \\
    $s = R_o - V1$ \\
    $u = f(s \cdot h)$ \\
    \uIf{$u < 0 \; OR \; u > 1$}{ return false }
    $q = s \times edge_1$ \tcp{Cross product}
    $v = f(R_d \cdot q)$ \\
    \uIf{$v < 0 \; OR \; v > 1$}{return false}
    $t = f(edge_2 \cdot q)$ \\
    \tcp{If triangle is behind ray origin}
    \uIf{$t < 0$}{return false}
    return true

    \caption{Ray Triangle Intersect}
\end{algorithm}

@@

3D models are stored in files in OBJ format with a list of vertices, list of normals, and list of faces where each face will have the indices of the vertices and normals. The OBJ files are loaded and parsed on the host and converted into a similar format designed for the kernels to use. A separate buffer is created to store the vertices, normals, and faces of every model and triangle which are passed to the kernels as kernel arguments. Models with many triangles will largely affect the intersection time as the rays will test intersection with every triangle without acceleration structures.

@@

The resulting light output from all of the secondary rays must be mixed to generate the final colour and output of the primary ray. Depending on the surface material properties (opacity, reflectance, etc), mixing the colours of the child rays will be different. The output result colour must be processed from the tail-end of the result stack as a parent ray must use the result of its child rays to produce its own output.

@@

The base colour output from when a ray intersects with an object is computed using the Phong model where the surface material's ambient, diffuse, and specular values are used. If the material is not reflective, completely opaque, and not in shadow, this will be the colour output for the ray. While not completely physically accurate, it provides a rough estimate of the shading and lighting of the object and is relatively fast \citep{bishop1986fast}. The Phong output will be combined with the other child ray outputs if the surface material allows it which may produce a better result. Algorithm \nameref{phongalgo} is used to compute the Blinn-phong colour output in the kernel \citep{blinn1977models} which is a faster Phong approximation algorithm and can be more accurate in some cases.

\begin{algorithm}
    \label{phongalgo}
    \SetAlgoLined
    \KwResult{Colour output computed with Phong}
    \KwData{$A, D, E, R, N, L$}
    \tcc{
        $A$: Ambient, $D$: Diffuse, $E$: Specular Exponent, $R$: Ray, $N$: Surface Normal, $L$: Light Direction
    }

    $I = N \cdot L$ \tcp{Light intensity}
    $C = I \times D$ \tcp{Diffuse colour}

    $H = normalize(L + R_d)$ \tcp{Blinn-Phong Half Vector}

    $I = (-N \cdot H)^E$ \tcp{Specular colour intensity}

    return $A + C + I$

    \caption{Compute Phong Colour}
\end{algorithm}

Figure \ref{phong_img} shows a non-reflective and opaque object in the scene with its colour output computed using the Phone model.

@@

The final output of a ray hitting a surface is simply a mixture of the reflection, shadow, and transmission. The transmission itself is a mix of the base material colour (Phong) and the light output of any rays that go through the object (refraction). When opacity is less than $1.0$ ($0.0$ is fully transparent and $1.0$ is fully opaque), refraction rays are traced and the result is linearly mixed with the base colour based on the opacity.

If a ray does not intersect with any object, the output becomes the background or sky colour dependent on the ray direction and there are no child rays created. Similarly, if the ray has reached the bounce limit, no child rays will be created for that ray.

Transmission for any ray is computed with $mix(refract, base, opacity)$ where the $mix$ function is a linear interpolation function, $refract$ is the colour output of the refract child ray, and $base$ is the base colour computed by the Phong model. If the ray has reached its bounce limit and does not have a refract child ray, the opacity is simply treated as $1.0$.

If a reflect child ray has been traced, the colour output of the reflect ray is mixed with the transmission based on a Fresnel factor which will be discussed in the next section.

Figure \ref{sphere_reflect} shows a fully opaque sphere with 100\% reflectance.

Figures \ref{sphere_refract_solid} and \ref{sphere_refract_hollow} show a fully transparent sphere with refractive index $1.157$ and $1.04$. For demonstration purposes, the spheres are shown with reflection omitted to only show the effects and results of refraction. Comparing the images, the amount the refractive index changes the direction of the refraction ray is clear.

Figure \ref{spheres_refractive_index} shows a lineup of transparent spheres with varying refractive indices.

@@

For accurate results, a transparent material such as glass should use a mix of both the transmission and reflection light however a linear mix is simply not accurate. A more physically accurate approach uses a factor based on the angle of incidence of the ray onto the surface. The transmission factor increases when the angle of incidence increases. With a fresnel factor of $F$, the surface can be coloured with $T(1 - F) + LF$ where $T$ is the transmission colour, $F$ is the fresnel factor (a value $0.0 \rightarrow 1.0$), and $L$ is the reflection colour. The algorithm \nameref{fresnelalgo} shows the implementation of how the fresnel factor is computed in the kernel. Figures \ref{fresnel_solid} and \ref{fresnel_hollow} show the same transparent refractive spheres with mixed reflection using the fresnel factor.

\begin{algorithm}
    \label{fresnelalgo}
    \SetAlgoLined
    \KwResult{Fresnel Factor}
    \KwData{$I, N, F, T$}
    \tcc{$I$: incidence direction vector, $N$: surface normal, $F$: refractive index of the ray medium, $T$: refractive index of the object medium}

    $cosI = I \cdot N$ \\
    $n = \frac{F}{T}$ \\
    $sinT = \sqrt{n * (1 - cosI^2)}$ \\
    $cosT = \sqrt(1 - sinT^2)$ \\
    $R_s = \frac{TcosI - FcosT}{TcosI + FcosT}$ \\
    $R_p = \frac{FcosI - TcosT}{FcosI + TcosT}$ \\
    return $\frac{R_s^2 + R_p^2}{2}$ \\

    \caption{Function for computing the fresnel factor}
\end{algorithm}

@@

The implementation of shadows reduce the light or colour intensity of the surface instead of the more physically accurate approach where light simply does not bounce on shadowed surfaces as much. If the shadow ray created with an origin at the surface intersects with an object, the surface is simple treated as in shadow. A surface is either in shadow or not in shadow and this boolean flag will determine whether the surface light intensity needs to be reduced or not. Surfaces in shadow are simply reduced by a fixed factor. This approach produces hard shadows as seen in figure \ref{hard_shadow}.

The performance impact is trivial however it does not produce results that are physically accurate or realistic. For softer shadows, multiple shadow rays can be traced to determine the strength of the shadow at a particular spot on the surface. The extra shadow rays have their origin offset by a small distance, still directed at the light source. Shadow rays near the edge will have a larger disproportion between shadow rays that intersect and rays that don't intersect. To make the surface shadow softer, the amount the intensity is reduced is lessened by the proportion of missed shadow rays to hit shadow rays. The more extra shadow rays are traced, the smoother the shadow.

When only using a few extra shadow rays, the soft shadow effect only gives it a slight anti-aliasing effect which may be beneficial and only impact performance slightly.

Figures \ref{softish_shadow} and \ref{soft_shadow} show the shadow with varying level for softness.

@@

For accurate compution of a ray's output, the output of its child rays must be resolved first. This can usually be achieved by running the colour compution procedure recursively on each of the rays children before computing the rays own output however recursion is not supported in the OpenCL kernel language. The alternative method implemented a workaround to recursion by using an array as a stack. Since dynamic arrays are also not supported, a static array with a maximum size limit was used with a variable to keep track of the stack head which represents the index of the static array which is free and a variable to keep track of the recursive traversal through the stack. The recursive function approach would have called the function recursively multiple times so the stack must represent a tree with $N$ children where $N$ is the number of child rays that can be generated off a single ray.

@@

The majority time taken during the rendering process is in the ray intersection tests. The naive approach is testing for intersection between a ray and every object in the scene however this is inefficient and some faster preliminary checks can be made to reduce the number of ray-primitive intersection tests. For example, a $1280 \times 720$ image with a single sphere taking up $200 \times 200$ pixels on the image need only perform ray-sphere intersection $40000$ times instead of the full $921600$ times if every primary ray were to perform the intersection test. However if the object were to take 100\% of the screen space, the preliminary checks may end up negatively impacting performance (although slight). Fortunately this is hardly ever the case during a typical scene.

Another way to improve render time is to optimise the intersection algorithms although this would have less impact than just reducing the number of tests. The acceleration structures: bounding volumes and grid, are used to reduce the number of intersection tests by omitting them from rays which can be determined to not intersect with the object without doing an intersection test. A perfect method would have rays only perform intersection tests with objects that it does intersect with however obviously this is farfetched and unreasonable but it is the target of acceleration structures.

@@

There are some simple checks that can be performed for objects with large number of triangles such as radial checks or AABB checks however some models and shapes may not make the best use of the area that is checked with these methods. For this reason, models are checked with a bounding volume. For spheres, before the intersection is performed, a simple dot product check is performed so the intersection test is skipped if the sphere is behind the ray. The preliminary checks do not have any improvement to performance of an individual ray if the ray does actually intersect with the object as the actual ray-primitive check will still be performed on top of the preliminary check if the preliminary check passes. Additionally, the preliminary checks, while fast, still take a non-zero amount of time and will still be slow when the number of objects is large.

@@

Bounding volumes work similar to AABB, checking for overlaps in the bounds of the $X, Y, Z$ axis. Bounding volumes aim to achieve closer bounds to the convex hull while still being computationally inexpensive. This is achieved by using the $X, Y, Z$ axis and an additional $4$ axis which cut the bounds in the diagonal. The axis are defined by: $(S, S ,S), (-S, S ,S), (-S, -S, S), (S, -S, S)$ where $S$ is defined by $\frac{\sqrt{3}}{3}$.
